Gammapy: A Python package for gamma-ray astronomy

### 2023-01-26 src/ms-review.tex
Address more of CNs comments

\abstract
	{

		Traditionally, \gammaray astronomy has been conducted
		by collaborations of scientists employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array (CTA), will be operated as open observatories.
		Alongside the data, the observatories will make available to the astronomical community
		software tools for their analysis. This necessity prompted the development, in the last decade,
		of open high-level astronomy software customised for high energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source python package for the analysis of \gammaray data,
		and illustrate the functionalities of the first long-term release of the software, version 1.0.
		Built on the modern python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different gamma-ray instruments. \gammapy complies with several well-established
		data conventions in high-energy astrophysics, providing serialised data products that are interoperable
		with other software.
	}
	{
		Starting from event list and instrument response functions,
		\gammapy provides pipelines for creating reduced data binned in energy and coordinates.
		To facilitate the rejection of the residual hadronic background, several techniques for background estimation
		are implemented in the package.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also implemented.
	}
	{
		After describing the structure of the package, we demonstrate the capabilities of \gammapy
		performing both traditional and novel analyses in gamma-ray astronomy with publicly available data:
		spectral and spectro-morphological analyses, estimation of the light curve of a source.
		The flexibility and power of the software are displayed in a final multi-instrument example,
		where data sets from different instruments, at different stages of data reduction,
		are simultaneously fitted with a flux model produced by another astrophysical software.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
\gammaray astronomy is a rather young field of research. The \gammaray~
range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their direction and energy.
VHE \gammaray astronomy covers the energy range from fews tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments: 
Imaging Atmospheric Cherenkov Telescopes (IACT) and 
Water Cherenkov observatories. IACTs obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

@@

Water Cherenkov observatories detect particles directly from the tail of the
shower when it reaches the ground. These instruments have very
large field-of-view, large duty-cycle but higher energy threshold and
usually have lower signal to noise ratios compared to IACTs \citep{2015CRPhy..16..610D}.

@@

%% Context : formats
%% CN: particle samplers -> WCD
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray-like events.
The lists of \gammaray events are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the observatory's specific instrument response functions (IRF).
Once the data is reduced to a list of events, the information is independent of
the data-reduction process, and, eventually, of the detection technique. This implies,
for example, that hihg-level data from IACT and particle samplers can be represented
with the same data model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). This proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system. %% where? in the data units of a FITS file.

%% Context: develoment of open software
Python has become extremely popular as a scientific programming language,
in particular, in the field of data sciences. This success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a 'glue' language between different programming languages and last but not least
the rich ecosystem of packages and its open and supportive community, from
core numerical analysis packages such as \numpy~\citep{numpy} and
\scipy~\citep{2020SciPy-NMeth}, visualization libraries, e.g. Matplotlib \citep{matplotlib},
to optimization packages such as \iminuit \citep{iminuit}.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project started in 2015 with the objective of building a flexible and
efficient software library for very high-energy \gammaray data analysis \citep{gammapy_2015}.
This is possible thanks to the definition of a standardised common data format such as the \gadf, and thanks to
an open development with a broad community of contributors. The relation of \gammapy to the
different instruments and the \gadf data format is illustrated in Figure~\ref{fig:big_picture}.

@@

\section{Gamma-ray Data Analysis}
% CN: gamma-like here, was \gammaray-like in the previous section
\begin{figure*}[h!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of gamma-like
        events on the left (DL3), to high-level scientific products products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data analysis process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This sequence, sometimes referred to as low-level analysis, is usually very specific to
a given observation technique and even to a given instrument.

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.
%% CN: I don't understand this: since the high-level methods are the same, then they "imply"
%% joint analyses?? Maybe they eventually allow or permit?

@@

In the most general way, we can write the expected number of detected events from the sky model $\Phi$ at measured position $p$ and energy $E$, for
a given set of parameters $\hat{\theta}$, as
\begin{align}
   N(p, E, \hat{\theta})\,{\rm d}p\,{\rm d}E = &t_{\rm obs} \int_{E_{\rm true}} \int_{p_{\rm true}}  R(p, E|p_{\rm true}, E_{\rm true}) \nonumber \\
   &\times \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta} ) {\rm d}E_{\rm true} {\rm d}p_{\rm true}
\end{align}
where $R(p, E| p_{\rm true}, E_{\rm true})$ is the instrument response and $t_{\rm obs}$ is the observation time

@@

\begin{align}
   R(p, E|p_{\rm true}, E_{\rm true}) = &A_{\rm eff}(p_{\rm true}, E_{\rm true}) \times \nonumber \\
    &PSF(p|p_{\rm true}, E_{\rm true}) \times \nonumber \\
    &E_{\rm disp}(E|p_{\rm true}, E_{\rm true})
\end{align}
where:
\begin{itemize}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments consider the probability density of the angular separation between true and reconstructed directions
  $\delta p = p_{\rm true} - p$, i.e. $PSF(\delta p|p_{\rm true}, E_{\rm true})$.
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider the probability density of the migration $\mu=\frac{E}{E_{\rm true}}$,
  i.e. $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$.
\end{itemize}

@@

In total, the expected number of events in a \gammaray observation is given by:
\begin{align}
  N(p, E, \hat{\theta})\ {\rm d}p\ {\rm d}E =  &E_{\rm disp} \star \left[ PSF \star \left( A_{\rm eff} \times t_{\rm obs} \times \Phi(\hat{\theta}) \right) \right]  \nonumber\\
                       & + Bkg(p, E) \times t_{\rm obs}
\end{align}
%% CN: Instead of rewriting this formula and introducing this new notation for the convolution,
%% can't we just say that, when predicting counts, $Bkg(p, E)$ has to be summed to the convolution
%% (or integral) in the previous equation?

@@

%% CN: Shouldn't we display the output of this snippet?
%% otherwise what's teh point of the `print`' functions?
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_data}
	\caption{
        Using \code{gammapy.data} to access DL3 level data with a \code{DataStore} object.
        Individual observations can be accessed by their unique integer observation id number.
        The actual events and instrument response functions can be accessed
        as attributes on the \code{Observation} object, such as \code{.events}
        or \code{.aeff} for the effective area information.
    }
	\label{fig*:minted:gp_data}
\end{figure}
%

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		Here we show example instrument response functions of some experiments and
        observatories for which \gammapy can analyse data. The \cta IRFs
        are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates.
        The point spread function shows the $68\%$
        containment radius of the PSF. The \fermi IRFs are from \textit{pass8}.
    }
	\label{fig:irfs}
\end{figure*}
%
%
%% CN: description of the HAWC IRF missing, what is the difference between the gray
%% and black effective areas? Where are they taken from?

>>>>

\abstract
	{
		Traditionally, \gammaray astronomy has been conducted
		by collaborations of scientists employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array (CTA), will be operated as open observatories.
		Alongside the data, the observatories will make available to the astronomical community
		software tools for their analysis. This necessity prompted the development, in the last decade,
		of open high-level astronomy software customised for high energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source python package for the analysis of astronomical \gammaray data,
		and illustrate the functionalities of the first long-term release of the software, version 1.0.
		Built on the modern python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different \gammaray instruments for a variety of analysis scenarios. \gammapy complies with several well-established
		data conventions in high-energy astrophysics, providing serialised data products that are interoperable
		with other software.
	}
	{
		Starting from event list and instrument response functions,
		\gammapy provides the funcionality for reducing data binned in energy and sky coordinates.
		To facilitate the rejection of the residual hadronic background, several techniques for background estimation
		are implemented in the package.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also supported.
	}
	{
		After describing the structure of the package, we show the capabilities of \gammapy
		in multiple traditional and novel \gammaray astronomy analysis scenarios using publicly available data
		such as spectral and spectro-morphological modelling as well as estimation of the spectral energy
		distribution (SED) and light curve of a source.
		The flexibility and power of the software are displayed in a final multi-instrument example,
		where datasets from different instruments, at different stages of data reduction,
		are simultaneously fitted with an astrophysical flux model produced by another astrophysical software.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
\gammaray astronomy is a rather young field of research. The \gammaray
range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from fews tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments: 
Imaging Atmospheric Cherenkov Telescopes (IACT) and 
Water Cherenkov Detectors (WCD). IACTs obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

@@

Water Cherenkov observatories detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV large duty-cycle but higher energy threshold and
usually have lower signal to noise ratios compared to IACTs \citep{2015CRPhy..16..610D}.

@@

%% Context : formats
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray-like events.
The lists of \gammaray events are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the observatory's specific instrument response functions (IRF).
Once the data is reduced to a list of events, the information is independent of
the data-reduction process, and, eventually, of the detection technique. This implies,
for example, that high-level data from IACTs and WCDs can be represented
with the same data model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). This proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
such as energy, incident direction and arrival time and a parametrisation of
the IRFs associated with the event list data. %% where? in the data units of a FITS file.

%% Context: develoment of open software
In the past deaced Python has become extremely popular as a scientific programming language,
in particular, in the field of data sciences. This success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a "glue" language between different programming languages and last but not least
the rich ecosystem of packages and its open and supportive community, from
core numerical analysis packages such as \numpy~\citep{numpy} and
\scipy~\citep{2020SciPy-NMeth}, visualization libraries, e.g. Matplotlib \citep{matplotlib},
to optimization packages such as \iminuit \citep{iminuit}.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project started based on the same idea as \astropuy with the objective of building a flexible and
efficient software library for very high-energy \gammaray data analysis \citep{gammapy_2015}.
This is possible thanks to the definition of a standardised common data format such as the \gadf, and thanks to
an open development with a broad community of contributors. The relation of \gammapy to the
different instruments and the \gadf data format is illustrated in Figure~\ref{fig:big_picture}.

@@

\section{Gamma-ray Data Analysis}
% CN: gamma-like here, was \gammaray-like in the previous section
\begin{figure*}[h!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of gamma-like
        events on the left (DL3), to high-level scientific products products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data analysis process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. The similarity in the high-level analysis would also allow
for combining data from multiple instruments, but could not be fully exploited, due to
a lack of common data formats and software tools.

@@

In the most general way, we can write the expected number of detected events from the sky model $\Phi$ at measured position $p$ and energy $E$, for
a given set of parameters $\hat{\theta}$, as:
%
\begin{equation}
	\begin{split}
   		N(p, E, \hat{\theta})\,{\rm d}p\,{\rm d}E = &t_{\rm obs} \int_{E_{\rm true}} \int_{p_{\rm true}}  R(p, E|p_{\rm true}, E_{\rm true}) \\
   		&\times \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta} ) {\rm d}E_{\rm true} {\rm d}p_{\rm true}
	\end{split}
\end{equation}

where $R(p, E| p_{\rm true}, E_{\rm true})$ is the instrument response and $t_{\rm obs}$ is the observation time

@@

\begin{equation}
	\begin{split}
   R(p, E|p_{\rm true}, E_{\rm true}) = &A_{\rm eff}(p_{\rm true}, E_{\rm true}) \times \\
    &PSF(p|p_{\rm true}, E_{\rm true}) \times \nonumber \\
    &E_{\rm disp}(E|p_{\rm true}, E_{\rm true})
	\end{split}
\end{equation}

where:
\begin{itemize}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments consider the probability density of the angular separation between true and reconstructed directions
  $\delta p = p_{\rm true} - p$, i.e. $PSF(\delta p|p_{\rm true}, E_{\rm true})$.
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider the probability density of the migration $\mu=\frac{E}{E_{\rm true}}$,
  i.e. $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$.
\end{itemize}

@@

In total, the expected number of events in a \gammaray observation is given by:
\begin{equation}
	\begin{split}
  N(p, E, \hat{\theta})\ {\rm d}p\ {\rm d}E =  &E_{\rm disp} \star \left[ PSF \star \left( A_{\rm eff} \times t_{\rm obs} \times \Phi(\hat{\theta}) \right) \right]\\
                       & + Bkg(p, E) \times t_{\rm obs}
	\end{split}
\end{equation}

%% CN: Instead of rewriting this formula and introducing this new notation for the convolution,
%% can't we just say that, when predicting counts, $Bkg(p, E)$ has to be summed to the convolution
%% (or integral) in the previous equation?

@@

%% CN: Shouldn't we display the output of this snippet?
%% otherwise what's the point of the `print`' functions?
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_data}
	\caption{
        Using \code{gammapy.data} to access DL3 level data with a \code{DataStore} object.
        Individual observations can be accessed by their unique integer observation id number.
        The actual events and instrument response functions can be accessed
        as attributes on the \code{Observation} object, such as \code{.events}
        or \code{.aeff} for the effective area information.
    }
	\label{fig*:minted:gp_data}
\end{figure}
%

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius o as a function of energy for the \cta, \hess
		and \fermi instruments. These IRFs do not correspond to the latest
		performance of the instruments, but still are representative of the 
		detector type and energy range. We also neede to rely on publicly available
		data. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of event data the \hawc collaboration. All 
		data is also availabe in the \code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%
